Factors affecting reliablity of state-space age-structured assessment models

### 2025-02-12 Project_0/manuscript/manuscript.tex
saving work on project_0 manuscript

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Brandon
Muffley\textsuperscript{5} \and Brian C.
Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{15 January, 2025}

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume
spawning occurs each year 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_maturity}).

Weight at age was generated with a von Bertalanffy growth function \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs (Miller and Brooks 2021). We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean log natural
mortality rate was \(\log(0.2)\).

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
1 (Figure \ref{om_mean_selectivity}). This selectivity is was used to
define \Fmsy for the Beverton-Holt stock recruitment parameters above.
We assumed a logistic-normal distribution for the age-composition
observations for the fleet.

@@

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 of each year) and one in the fall (0.75 of each year).
Catchability of both surveys are assumed to be 0.1. Like the fishing
fleet, we assumed logistic selectivity for both indices with
\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution for the
age-composition observations.

@@

All R+M OMs treat natural mortality constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean that is constant across ages
\(\mu_{M} = \log(0.2)\) and any annual random effects marginally
distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, R+M). The marginal standard deviations we assumed for log
natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
factorial combination of these process error assumptions and fishing
history (2 levels) and observation error (2 levels) scenarios described
above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

For each data set simulated from an OM 20 EMs were fit. A total of 32
different EMs were fit across all OMs where the subset of 20 depended on
the source of process error (Table \ref{em_table}). The first 20 EMs in
Table \ref{em_table} were fit to simulate data sets from R and R+S OMs.
EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from
R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to
simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32 in
Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
observation error variance of aggregate catch and indices were all
assumed known at the true values.

maturity, weight at age, index and catch CVs, assumed known in all EMs

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit. We summarized 5
alternative categories of convergence.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Fit complete: Did the optimization routine (stats::nlminb) complete
  without error?
\item
  No flag: Did the stats::nlminb convergence flag = 0 indicate
  successful convergence?
\item
  Good gradient: Was the maximum absolute value of the gradient of the
  log-likelihood \textless{} \(1\times10^{-6}\)?
\item
  No SE NAs: Did TMB::sdreport provide non-NA values for all fixed
  effects standard errors?
\item
  No big SEs: Did TMB::sdreport provide all standard errors \textless{}
  100?
\end{enumerate}

The first convergence criterion assesses whether the model crashes. The
third is a measure that assesses how flat the likelihood is at the
optimized point of the likelihood surface. The fourth and fifth criteria
are specific to the hessian-based standard error reporting provided by
TMB. the TMB::sdreport function will sometimes return standard error
estimates even when the calculated hessian is not invertible (fourth
criterion). It will also provide large standard errors for some
parameters that are not estimated well or are near bounds on the
transformed scale that is of primary interest. For example, variance
parameters for random effects can often be estimated near 0 when the fit
suggests no variation in the random effects or, equivalently, that the
model without random effects is better. Note that Types 2 through 5 also
require type 1, and type 5 requires type 4. We used the Clopper-Pearson
exact method for constructing 95\% confidence intervals of the
probabilities of convergence (Clopper and Pearson 1934; Thulin 2014).

@@

We estimated the probability of selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
operating model, we compared AIC for EMs that all made the same
assumptions about median/constant natural mortality (known or estimated)
and stock recruitment model (Beverton-Holt or none).

We also estimated the probability of correctly selecting between models
with Beverton-Holt stock recruit function assumed and models without the
stock-recruit function (null model). We made these comparisons between
models that otherwise assume the same process error structure as the
operating model and both of the compared models either estimate
median/constant natural mortality or assume it is known. Our preliminary
inspections of the proportions of simulations where the correct
recruitment model was chosen for a given set of OM factors indicated
generally poor performance of AIC. Furthermore, it has been shown that
estimation of stock-recruit parameters requires a wide range of SSB
(References). Therefore, we fit logistic regression models to the
indicator of Beverton-Holt models having lower AIC as a function of the
log-standard deviation of the true log(SSB) (similar to the log of the
coefficient of variation for SSB) for each simulation.

@@

We estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] of annual SSB and fully-selected fishing mortality for each EM and
constructed 95\% confidence intervals for the median relative bias using
the binomial distribution approach as in Miller and Hyun (2018) and
Stock and Miller (2021). We also estimated bias of stock-recruit
parameters for EMs that assumed the Beverton-Holt stock recruit function
and the mean natural mortality rate parameter when estimated.

@@

We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
and recruitment for each EM. We estimated 7 peels for each EM. We
calculated median 95\% confidence intervals for Mohn's \(\rho\) using
the same methods as that for relative bias.

Similar to the other results, retrospective results only condition on
whether all of the peels of a given estimating model completed the
optimization process without failure. We did not condition on
convergence as defined by a gradient threshold or invertibity of the
hessian because the optimized model can provide reliable estimation of
SSB, F, M, and stock recruit parameters whether or not the model was
able to estimate non-zero random effects.

@@

\hypertarget{r-rs-operating-models}{%
\subsubsection*{R, R+S operating models}\label{r-rs-operating-models}}
\addcontentsline{toc}{subsubsection}{R, R+S operating models}

R+S EMs fit to R OMs exhibited poor convergence for types 3, 4, and 5
regardless of whether M or a stock-recruit relationship was estimated
(Figures \ref{naa_om_em_R_MF_convergence} to
\ref{naa_om_em_BH_ME_convergence}). R EMs exhibited high convergence
rates for all convergence types for both R and R+S OMs. Convergence
rates were high for all convergence types for all EMs in R+S OMs with
lower uncertainty in observations. Convergence for most convergence
types generally declines most EMs when mean-log natural mortality rate
is estimated and/or a Beverton-Holt stock recruit relationship is
estimated even when the process error assumptions of the estimation and
OMs match. There can be relatively high convergence probability of type
4 (an invertible hessian) for EMs that do not have the correct process
error assumed, but type 5 convergence (no very large standard errors
estimated from hessian) for these EMs is typically much lower.

\hypertarget{rm-operating-models}{%
\subsubsection*{R+M operating models}\label{rm-operating-models}}
\addcontentsline{toc}{subsubsection}{R+M operating models}

R+S EMs generally converged less reliably across convergence types than
other EMs when fit to data generated from R+M OMs (Figures
\ref{M_om_em_R_MF_convergence} to \ref{M_om_em_BH_ME_convergence}).
However, using convergence type 5, all EMs converged poorly when OMs had
low variability in natural mortality process errors or higher
observation error uncertainty. Using convergence type 5, R+M EMs
generally converged better than other EMs when observation error
uncertainty was low and natural mortality process errors were more
variable. Like R+S OMs, the probability of convergence generally
declined for all EMs when mean log-natural mortality and/or a
stock-recruit relationship was estimated.

\hypertarget{rsel-operating-models}{%
\subsubsection*{R+Sel operating models}\label{rsel-operating-models}}
\addcontentsline{toc}{subsubsection}{R+Sel operating models}

R+S EMs in particular converged poorly when R+Sel OMs had less
variability in selectivity process errors or higher observation error
uncertainty (Figures \ref{Sel_om_em_R_MF_convergence} to
\ref{Sel_om_em_BH_ME_convergence}). R+Sel EMs generally converged better
than other EMs for OMs with greater variability in process errors, lower
observation error, and contrast in fishing pressure regardless of
whether mean log natural mortality or a stock recruit relationship was
estimated.

\hypertarget{rq-operating-models}{%
\subsubsection*{R+q operating models}\label{rq-operating-models}}
\addcontentsline{toc}{subsubsection}{R+q operating models}

Convergence of R+q EMs is generally better than that of other EMs for
all convergence types when R+q OMs assume contrast in fishing history
overFigures (\ref{q_om_em_R_MF_convergence} to
\ref{q_om_em_BH_ME_convergence}). Convergence of R+S EMs is generally
worse than that of other EMs across all OMs whether or not mean
log-natural mortality or a stock recruit relationship is estimated.
Again, convergence probablity generally declines for all EMs when mean
log-natural mortality or a stock recruit relationship is estimated.

@@

\hypertarget{r-rs-operating-models-1}{%
\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-1}}
\addcontentsline{toc}{subsubsection}{R, R+S operating models}

Marginal AIC accurately determines the correct process error assumptions
in EMs when data are generated from R and R+S OMs, regardless of whether
mean log-natural mortality or a stock recruit relationship is estimated
(Figures \ref{naa_om_proportion_best_aic_R_MF} to
\ref{naa_om_proportion_best_aic_SR_ME}). Adding estimation of mean
log-natural mortality or a stock recruit relationship separately has a
negligible effect on the accuracy of determining the correct process
error assumption. When both are estimated, there is a noticeable
reduction in accuracy when OMs have a constant fishing history,
observation error is low and largest variability in recruitment process
errors.

\hypertarget{rm-operating-models-1}{%
\subsubsection*{R+M operating models}\label{rm-operating-models-1}}
\addcontentsline{toc}{subsubsection}{R+M operating models}

Marginal AIC only accurately determined the correct process error model
and correlation structure when observation error was low and variability
in natural mortality process errors was high (Figures
\ref{M_om_proportion_best_aic_R_MF} to
\ref{M_om_proportion_best_aic_SR_ME}). Estimating the mean natural
mortality rate reduced the accuracy of AIC for OMs that assumed natural
mortality process errors were independent. For OMs with poor accuracy,
AIC most frequently selected EMs with process errors in catchability or
selectivity. Selection of R+S EMs was typically unlikely.

\hypertarget{rsel-operating-models-1}{%
\subsubsection*{R+Sel operating models}\label{rsel-operating-models-1}}
\addcontentsline{toc}{subsubsection}{R+Sel operating models}

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+Sel OMs with low observation error
(Figures \ref{Sel_om_proportion_best_aic_R_MF} to
\ref{Sel_om_proportion_best_aic_SR_ME}). When there was low variability
in selectivity process errors and high observation error, R+q or R+S EMs
were more likely to have the best AIC. Whether mean log-natural morality
or stock recruit relationships were estimated appeared to have little
effect on the performance of marginal AIC.

\hypertarget{rq-operating-models-1}{%
\subsubsection*{R+q operating models}\label{rq-operating-models-1}}
\addcontentsline{toc}{subsubsection}{R+q operating models}

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+q OMs with high variability in
catchability process errors (Figures \ref{q_om_proportion_best_aic_R_MF}
to \ref{q_om_proportion_best_aic_SR_ME}). The worst accuracy occurred
for OMs with low variability in catchability process errors and high
observation error. However, in these OMs, the marginal AIC accurately
determined the correct source of process error (but not correlation
structure) except when EMs estimated both mean log-natural morality and
the stock recruit relationships and OMs assumed a constant fishing
pressure.

@@

Our comparisons of model performance condition on assuming the true
process error configuration is known (EM and OM process error
assumptions match). Broadly, we found generally poor accuracy of AIC in
selecting models assuming a Beverton-Holt stock recruit function over
the null model without an assumed stock-recruit relationship for all OMs
(Tables \ref{naa_om_em_R_BH_aic_table} to \ref{q_om_em_R_BH_aic_table}.
However, we also found AIC to more accurately determine a stock-recruit
relationship when there was greater variation in spawning biomass
generated in the simulated populations.

\hypertarget{r-rs-operating-models-2}{%
\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-2}}
\addcontentsline{toc}{subsubsection}{R, R+S operating models}

Among R and R+S OMs, AIC accuracy for including the B-H stock-recruit
relationship was estimated to be highest for R OMs and EMs and lowest
variance of recruitment process errors (Figures
\ref{naa_om_MF_BH_glm_AIC_plots} and \ref{naa_om_ME_BH_glm_AIC_plots}).
FOr R+S OMs and holding variability in SSB constant, accuracy of
stock-recruit model selection declined with greater variance of
recruitment and survival process errors. Simultaneously estimating mean
log-natural mortality has a small negative effect on AIC accuracy, but
there is an improvement in AIC accuracy with lower observation error in
indices and age composition and lower variation of survival process
errors.

\hypertarget{rm-operating-models-2}{%
\subsubsection*{R+M operating models}\label{rm-operating-models-2}}
\addcontentsline{toc}{subsubsection}{R+M operating models}

For R+M OMs and EMs there appeared to be little effect of correlation in
the natural mortality process errors or estimating the mean log-natural
mortality on accuracy of AIC in determining a stock-recruit function,
holding variability in SSB constant (Figures
\ref{M_om_MF_BH_glm_AIC_plots} and \ref{M_om_ME_BH_glm_AIC_plots}).
However, there was an increase in accuracy with lower observation error
for OMs with higher variability in natural mortality process errors
whether the mean natural mortality rate parameter was estimated or not.

\hypertarget{rsel-operating-models-2}{%
\subsubsection*{R+Sel operating models}\label{rsel-operating-models-2}}
\addcontentsline{toc}{subsubsection}{R+Sel operating models}

The range of variation in SSB across simulated populations was lower for
R+Sel OMs than R+S or R+M OMs. OMs with contrast in fishing pressure
over time is the primary mechanism creating variation in SSB. However,
there appears to be little effect of variability or correlation of
selectivity process errors or whether mean log-natural mortality was
estimated on the accuracy of AIC in selecting the stock recruit
relationship (Figures \ref{Sel_om_MF_BH_glm_AIC_plots} and
\ref{Sel_om_ME_BH_glm_AIC_plots}). However, there was an increase in AIC
accuracy with lower observation error for OMs with independent process
errors whether the mean natural mortality rate parameter was estimated
or not.

\hypertarget{rq-operating-models-2}{%
\subsubsection*{R+q operating models}\label{rq-operating-models-2}}
\addcontentsline{toc}{subsubsection}{R+q operating models}

Like the R+Sel OMs, the range of variation in SSB across simulated
populations for R+q OMs is only increased by allowing contrast in
fishing pressure over time. There appears to be a slight decrease in
accuracy in AIC when mean log-natural mortality is estimated and there
is a more noticeable increase in accuracy for OMs with lower observation
error in those cases. (Figures \ref{q_om_MF_BH_glm_AIC_plots} and
\ref{q_om_ME_BH_glm_AIC_plots}). There was also an increase in AIC
accuracy with lower observation error for when the mean natural
mortality rate parameter was estimated.

@@

\hypertarget{r-rs-operating-models-3}{%
\subsubsection*{R, R+S operating models}\label{r-rs-operating-models-3}}
\addcontentsline{toc}{subsubsection}{R, R+S operating models}

When mean log-natural mortality is known, all estimating models
regardless of process error assumptions produce unbiased estimates of
SSB over the entire time period for R OMs (Figure
\ref{naa_om_em_R_MF_relbias_ssb}). For R+S OMs, only R+S EMs produce
unbiased estimation of SSB and bias resulting from other EMs increased
with larger variability in survival random effects. Estimating mean
log-natural mortality generally produced greater variability in relative
errors of SSB estimates and a change in sign of bias for EMs assuming
process errors other than R and R+S (Figure
\ref{naa_om_em_R_ME_relbias_ssb}). Estimating a stock-recruit
relationship had little discernable effect on bias of SSB estimation
(Figures \ref{naa_om_em_BH_MF_relbias_ssb}), but in combination with
estimating mean log-natural mortality resulted in large bias of R+S EMs
for R OMs with high observation error, constant fishing pressure and
large variability in recruitment process errors (Figure
\ref{naa_om_em_BH_ME_relbias_ssb}).

\hypertarget{rm-operating-models-3}{%
\subsubsection*{R+M operating models}\label{rm-operating-models-3}}
\addcontentsline{toc}{subsubsection}{R+M operating models}

>>>>

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Brandon
Muffley\textsuperscript{5} \and Brian C.
Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{12 February, 2025}

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume
spawning occurs each year 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_inputs_fig}, top left).

Weight at age was generated with a von Bertalanffy growth function \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_inputs_fig}, top right).

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs (Miller and Brooks 2021). We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom
right). For OMs without process errors on natural mortality we assumed
the rate was assumed 0.2. For OMs with process errors on natural
mortality the median natural mortality rate was 0.2.

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
1 (Figure \ref{om_inputs_fig}, bottom left). This selectivity is was
used to define \Fmsy for the Beverton-Holt stock recruitment parameters
above. We assumed a logistic-normal distribution with no correlation on
the multivariate normal scale for the age-composition observations for
the fleet.

@@

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 of each year) and one in the fall (0.75 of each year).
Catchability of both surveys are assumed to be 0.1. Like the fishing
fleet, we assumed logistic selectivity for both indices with
\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution with no
correlation on the multivariate normal scale for the age-composition
observations.

@@

All R+M OMs treat natural mortality constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean log-natural mortality
parameter that is constant across ages \(\mu_{M} = \log(0.2)\) and any
annual random effects marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, R+M). The marginal standard deviations we assumed for log
natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
factorial combination of these process error assumptions and fishing
history (2 levels) and observation error (2 levels) scenarios described
above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

For each data set simulated from an OM 20 EMs were fit. A total of 32
different EMs were fit across OMs where the subset of 20 depended on the
source of process error in the OM (Table \ref{em_table}). The EMs make
different assumptions about the source of process error (R+S, R+M,
R+Sel, R+q) and whether there is temporal autocorrelation or not, and
whether a Beverton-Holt stock recruit function is estimated and the
natural mortality rate (\(\mu_M\), the constant or mean on log scale for
R+M EMs) is estimated. For simplicity we refer to the derived estimate
\(e^{\mu_M}\) as the median natural mortality rate regardless of whether
natural mortality random effects are estimated in the EM. The first 20
EMs in Table \ref{em_table} were fit to simulate data sets from R and
R+S OMs. EMs 5 to 24 in Table \ref{em_table} were fit to simulate data
sets from R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were
fit to simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32
in Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
maturity at age, weight at age for catch and SSB, and observation error
variance of aggregate catch and indices were all assumed known at the
true values.

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021), but given the importance of estimates of uncertainty when using
assessment models in management, we estimated probablity of convergence
as measured by occurrence of a positive-definite hessian matrix at the
optimized negative log-likelihood that could be inverted. However, we
also provided results in the Supplementary Materials for the maximum of
the absolute values among all gradients for all fits of a given EM to
all simulated data sets from a given OM that produced hessian-based
standard errors for all estimated fixed effects. This provides an
indication of how poor the calculated gradients can be, but still
presumably converged adequately enough for parameter inferences.

@@

We estimated the probability of selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
operating model, we compared AIC for EMs that all made the same
assumptions about median natural mortality (known or estimated) and
stock recruitment model (Beverton-Holt or none).

We also estimated the probability of correctly selecting between models
with Beverton-Holt stock recruit function assumed and models without the
stock-recruit function (null model). We made these comparisons between
models that otherwise assume the same process error structure as the
operating model and both of the compared models either estimate median
natural mortality or assume it is known. Contrast in fishing pressure
and time series with recruitment at low stock size has been shown to
improve estimability of stock-recruitment parameters (Magnusson and
Hilborn 2007; Conn et al. 2010). Our preliminary inspections of the
proportions of simulations where the correct recruitment model was
chosen for a given set of OM factors (including contrast in fishing
pressure) indicated generally poor performance of AIC. Therefore, we fit
logistic regression models to the indicator of Beverton-Holt models
having lower AIC as a function of the log-standard deviation of the true
log(SSB) (similar to the log of the coefficient of variation for SSB)
since simulations with realized SSB producing low and high recruitments
would have larger variation in realized SSB.

@@

For a given model attribute we calculated the relative error \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] from fitting a given estimating model to simulated data set \(i\)
configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
are the estimated and true values for simulation \(i\). We estimated
bias as the median of the relative errors across all simulations for a
given OM and EM combination. We constructed 95\% confidence intervals
for the median relative bias using the binomial distribution approach as
in Miller and Hyun (2018) and Stock and Miller (2021). We present
results for terminal years SSB and recruitment, Beverton-Holt stock
recruit parameters (\(a\) and \(b\)), and median natural mortality rate.
Results for terminal year fishing mortality are strongly negatively
correlated with those for SSB and provided in the supplementary
materials.

@@

We estimated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
and recruitment for each EM (Mohn 1999). We estimated 7 peels for each
EM. We calculated median 95\% confidence intervals for Mohn's \(\rho\)
using the same methods as that for relative bias. Similar to the other
results, retrospective results only condition on whether all of the
peels of a given estimating model completed the optimization process
without failure. We did not condition on convergence as defined by a
gradient threshold or invertibity of the hessian because the optimized
model can provide reliable estimation of SSB, fishing mortality rates,
and recruitment whether or not the model was able to estimate non-zero
random effects.

@@

For R and R+S OMs, convergence generally declines for most EMs when the
median natural mortality rate is estimated and/or the Beverton-Holt
stock recruit relationship is estimated even when the process error
assumptions of the EMs and OMs match (Figure
\ref{hessian_SE_convergence}, A). When there was high observation error
and constant fishing pressure (change from \(2.5F_{\text{MSY}}\) to
\(F_{\text{MSY}}\)), convergence of all EM process error configurations
other than R EMs was poor for R OMs (\(\sigma_{2+} = 0\)) regardless of
whether median natural mortality and stock-recruit relationships were
estimated. Convergence of R EMs was high for all R and R+S OMs except
when there was high observation error and constant fishing pressure and
median natural mortality and stock-recruit relationships were estimated.
R+S EMs fit to R OMs exhibited poor convergence regardless of whether
natural mortality or a stock-recruit relationship was estimated. R+S EMs
fit to R+S OMs had highest convergence rates when there was contrast in
fishing pressure and low observation error. Convergence rates were high
for all EMs when fit to data from R+S OMs with lower observation
uncertainty except those where median natural mortality and/or
stock-recruit relationships were estimated.

Convergence of all EMs fitted to R+M OMs was best when the OMs had
higher natural mortality process error variability, low observation
error, and contrast in fishing pressure (Figure
\ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation
of process errors had poor convergence for R+M OMs when there was low
natural mortality process error variability whether or not there was
autocorrelation of the simulated process errors. R+S EMs fitted to data
generated from R+M OMs always converged poorly whether median natural
mortality and the Beverton-Holt stock-recruit relationship were
estimated.

R+S EMs, in particular, converged poorly when fit to data generated from
R+Sel OMs with lower selectivity process error variability or higher
observation uncertainty (Figure \ref{hessian_SE_convergence}, C). R+Sel
EMs generally converged better than other EMs for R+Sel OMs with higher
process error variability, lower observation uncertainty, and contrast
in fishing pressure regardless of whether median natural mortality or a
stock recruit relationship was estimated.

For R+q OMs, convergence of R+q EMs is generally better than that of
other EMs when there is contrast in fishing history (Figure
(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs is generally
worse than that of all other EMs across all OMs whether or not median
natural mortality or a stock recruit relationship is estimated. Again,
convergence probability generally declines for all EMs when median
natural mortality or a stock recruit relationship is estimated.

We found a wide range of maximum absolute values of gradients for models
that converged as defined here (Figure \ref{hess_grad}). The largest
value observed for a given EM and OM combination was typically
\(<10^{-3}\), but many converged models had values greater than 1. For
many OMs, EMs that assumed the correct process error type and did not
estimate median natural mortality or the Beverton-Holt stock-recruit
relationship produced the lowest gradient values.

@@

Marginal AIC accurately determines the correct process error assumptions
in EMs when data are generated from R and R+S OMs, regardless of whether
median natural mortality or a stock recruit relationship is estimated
(Figure \ref{pe_aic}, A). Adding estimation of median natural mortality
or a stock recruit relationship separately has a negligible effect on
the accuracy of determining the correct process error assumption. When
both are estimated, there is a noticeable reduction in accuracy when OMs
have a constant fishing history, observation uncertainty is low and
larger variability in recruitment process errors.

For R+M OMs, marginal AIC only accurately determined the correct process
error model and correlation structure when observation error was low and
variability in natural mortality process errors was high (Figures
\ref{pe_aic}, B). Estimating the median natural mortality rate reduced
the accuracy of AIC for OMs that assumed natural mortality process
errors were independent. For OMs with poor accuracy, AIC most frequently
selected EMs with process errors in catchability (R+q) or selectivity
(R+Sel). Selection of R+S EMs was typically unlikely.

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+Sel OMs with low observation
uncertainty (Figures \ref{pe_aic}, C). When there was low variability in
selectivity process errors and high observation uncertainty, R+q or R+S
EMs were more likely to have the best AIC. Whether median natural
morality or stock recruit relationships were estimated appeared to have
little effect on the performance of AIC.

Marginal AIC most accurately determined the correct source of process
error and correlation structure for R+q OMs with high variability in
catchability process errors (Figures \ref{pe_aic},D). The worst accuracy
occurred for OMs with low variability in catchability process errors and
high observation uncertainty. However, in these OMs, the marginal AIC
accurately determined the correct source of process error (but not
correlation structure) except when EMs estimated both median natural
morality and the stock recruit relationships and OMs assumed a constant
fishing pressure.

@@

Our comparisons of model performance condition on assuming the true
process error configuration is known (EM and OM process error types
match). Broadly, we found generally poor accuracy of AIC in selecting
models assuming a Beverton-Holt stock recruit function over the null
model without an assumed stock-recruit relationship for all OMs.
However, we also found increased accuracy of AIC in determining the
Beverton-Holt stock-recruit relationship with greater variation in
spawning biomass generated in the simulated populations for nearly every
OM (Figure \ref{sr_aic}).

With R and R+S process error assumptions, probability of AIC accuracy
for the B-H stock-recruit relationship as a function of SSB variability
were greatest for OMs where there was greater process variability in
survival, lower process variability in recruitment and contrast in
fishing pressure (Figure \ref{sr_aic}, A). However, the largest
variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs
with larger recruitment variability (\(\sigma_R = 1.5\); Figure
\ref{sr_aic}, A, right column group). The weakest overall accuracy of
AIC for the Beverton-Holt stock-recruit function was when there was
constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth
rows). The largest effect of simultaneously estimating median natural
mortality was when there was OMs had high observation uncertainty and
constant fishing pressure (Figure \ref{sr_aic}, A, second row).

For R+M OMs, probability of AIC accuracy for the Beverton-Holt
stock-recruit relationship was greatest for OMs where there was contrast
in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in
SSB occurred in OMs with larger natural mortality variability and
temporal AR1 correlation of the annual random effects
(\(\sigma_M = 1.5\), \(\rho_M = 0.9\); Figure \ref{sr_aic}, B,
right-most column) where AIC accuracy was also high even when there was
constant fishing pressure. Interestingly, there seemed to be little
difference in AIC accuracy whether there was lower or higher observation
uncertainty or whether the EM simultaneously estimated median natural
mortality.

For R+Sel OMs, contrast in fishing pressure over time is the primary
mechanism creating variation in SSB and these are the OMs where AIC
accuracy for the Beverton-Holt stock recruit relationship was greatest
(Figure \ref{sr_aic}, C, first and third rows). There appears to be
little effect of variability or correlation of selectivity process
errors or whether median natural mortality was estimated on AIC
accuracy.

Like the R+Sel OMs, the greatest accuracy for AIC in selecting the
Beverton-Holt stock-recruit relationship occurred for R+q OMs where
there was contrast in fishing pressure over time which is also where
there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first
and third rows). There were some differences in effects of SSB variation
on AIC accuracy when OMs had high observation uncertainty and constant
fishing pressure. Across all of the OMs in Figure \ref{sr_aic}, the only
estimated decline in AIC accuracy with SSB variation occurred when R+q
OMs had high observation uncertainty and constant fishing pressure and
greater variation in catchability process errors that also had AR1
autocorrelation.

@@

\hypertarget{spawning-stock-biomass-and-recruitment}{%
\subsubsection*{Spawning stock biomass and
recruitment}\label{spawning-stock-biomass-and-recruitment}}
\addcontentsline{toc}{subsubsection}{Spawning stock biomass and
recruitment}

For R OMs (\(\sigma_{2+} = 0\)), there was no indication of bias (95\%
confidence intervals included 0) in terminal year SSB for any of the
estimating models regardless of process error assumptions, except when
no SR assumption was made, recruitment variability was low, and there
was contrast in fishing mortality and high observation uncertainty
(Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB
estimates were highly variable when median natural mortality was
estimated and there was constant fishing pressure and high observation
uncertainty (Figure \ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumption generally
produced unbaised estimation of terminal SSB except when median natural
mortality was estimated and there was high observation uncertainty. In
R+S OMs with low observation uncertainty, EMs with incorrect process
error assumptions typically provided biased estimation of terminal year
SSB. Estimating the Beverton-Holt stock-recruit relationship had little
discernible effect on bias of terminal year SSB estimation whereas
estimating median M tended to produce more variability in errors in
terminal SSB estimation similar to R OMs.
